---
title: "Data Visualization with ggplot"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`" 
output: 
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r setting up}
#install.packages("ggplot2")
library(ggplot2)
library(datasets)
```

First, you need to tell ggplot what dataset to use. This is done using the ggplot(df) function, where df is a dataframe that contains all features needed to make the plot. This is the most basic step. Unlike base graphics, ggplot doesn’t take vectors as arguments.
Optionally you can add whatever aesthetics you want to apply to your ggplot (inside aes() argument) - such as X and Y axis by specifying the respective variables from the dataset. The variable based on which the color, size, shape and stroke should change can also be specified here itself. The aesthetics specified here will be inherited by all the geom layers you will add subsequently.

If you intend to add more layers later on, may be a bar chart on top of a line graph, you can specify the respective aesthetics when you add those layers.

Below, I show few examples of how to setup ggplot using in the diamonds dataset that comes with ggplot2 itself. However, no plot will be printed until you add the geom layers.
```{r}
library(ggplot2)
data("midwest", package = "ggplot2")  # load the data
dim(midwest)
names(midwest)


# Init Ggplot
ggplot(midwest, aes(x=area, y=poptotal))  # area and poptotal are columns in 'midwest'
#A blank ggplot is drawn. Even though the x and y are specified, there are no points or lines in it. This is because, ggplot doesn’t assume that you meant a scatterplot or a line chart to be drawn. I have only told ggplot what dataset to use and what columns should be used for X and Y axis. I haven’t explicitly asked it to draw any points.

#Also note that aes() function is used to specify the X and Y axes. That’s because, any information that is part of the source dataframe has to be specified inside the aes() function.
```
## How to Make a Simple Scatterplot
Let’s make a scatterplot on top of the blank ggplot by adding points using a geom layer called geom_point.
```{r}
ggplot(midwest, aes(x=area, y=poptotal)) + geom_point()

```

## Adjusting the X and Y axis limits
The X and Y axis limits can be controlled in 2 ways.
This can be done by xlim() and ylim(). You can pass a numeric vector of length 2 (with max and min values) or just the max and min values itself.

```{r}
 g <- ggplot(midwest, aes(x=area, y=poptotal)) + geom_point() + geom_smooth(method="lm")  # set se=FALSE to turnoff confidence bands

# Delete the points outside the limits
g + xlim(c(0, 0.1)) + ylim(c(0, 1000000))   # deletes points
# g + xlim(0, 0.1) + ylim(0, 1000000)   # deletes points

```
## How to Change the Title and Axis Labels
I have stored this as g1. Let’s add the plot title and labels for X and Y axis. This can be done in one go using the labs() function with title, x and y arguments. Another option is to use the ggtitle(), xlab() and ylab().

```{r}
g <- ggplot(midwest, aes(x=area, y=poptotal)) + geom_point() + geom_smooth(method="lm")  # set se=FALSE to turnoff confidence bands

g1 <- g + coord_cartesian(xlim=c(0,0.1), ylim=c(0, 1000000))  # zooms in

# Add Title and Labels
g1 + labs(title="Area Vs Population", subtitle="From midwest dataset", y="Population", x="Area", caption="Midwest Demographics")

# or

g1 + ggtitle("Area Vs Population", subtitle="From midwest dataset") + xlab("Area") + ylab("Population")


```

##  How to Change the Color and Size of Points
How to Change the Color and Size To Static?
We can change the aesthetics of a geom layer by modifying the respective geoms. Let’s change the color of the points and the line to a static value.

```{r}
ggplot(midwest, aes(x=area, y=poptotal)) + 
  geom_point(col="steelblue", size=1) +   # Set static color and size for points
  geom_smooth(method="lm", col="firebrick") +  # change the color of line
  coord_cartesian(xlim=c(0, 0.1), ylim=c(0, 1000000)) + 
  labs(title="Area Vs Population", subtitle="From midwest dataset", y="Population", x="Area", caption="Midwest Demographics")

```
## How to Change the Color To Reflect Categories in Another Column?
Suppose if we want the color to change based on another column in the source dataset (midwest), it must be specified inside the aes() function.
```{r}
gg <- ggplot(midwest, aes(x=area, y=poptotal)) + 
  geom_point(aes(col=state), size=3) +  # Set color to vary based on state categories.
  geom_smooth(method="lm", col="firebrick", size=2) + 
  coord_cartesian(xlim=c(0, 0.1), ylim=c(0, 1000000)) + 
  labs(title="Area Vs Population", subtitle="From midwest dataset", y="Population", x="Area", caption="Midwest Demographics")
plot(gg)

```
## Theme
```{r}
gg + theme_bw() + labs(subtitle="BW Theme")
gg + theme_classic() + labs(subtitle="Classic Theme")

gg <- ggplot(midwest, aes(x=area, y=poptotal)) + 
  geom_point(aes(col=state, size=popdensity)) + 
  geom_smooth(method="loess", se=F) + xlim(c(0, 0.1)) + ylim(c(0, 500000)) + 
  labs(title="Area Vs Population", y="Population", x="Area", caption="Source: midwest")

plot(gg)

```

## How to Change Legend Labels and Point Colors for Categories
This can be done using the respective scale_aesthetic_manual() function. The new legend labels are supplied as a character vector to the labels argument. If you want to change the color of the categories, it can be assigned to the values argument as shown in below example.

```{r}
gg <- ggplot(midwest, aes(x=area, y=poptotal)) + 
  geom_point(aes(col=state, size=popdensity)) + 
  geom_smooth(method="loess", se=F) + xlim(c(0, 0.1)) + ylim(c(0, 500000)) + 
  labs(title="Area Vs Population", y="Population", x="Area", caption="Source: midwest")

gg + scale_color_manual(name="State", 
                        labels = c("Illinois", 
                                   "Indiana", 
                                   "Michigan", 
                                   "Ohio", 
                                   "Wisconsin"), 
                        values = c("IL"="blue", 
                                   "IN"="red", 
                                   "MI"="green", 
                                   "OH"="brown", 
                                   "WI"="orange"))

```

## Adding Text, Label and Annotation
How to Add Text and Label around the Points
Let’s try adding some text. We will add text to only those counties that have population greater than 300K. In order to achieve this, I create another subsetted dataframe (midwest_sub) that contains only the counties that qualifies the said condition.

Then, draw the geom_text and geom_label with this new dataframe as the data source. This will ensure that labels (geom_label) are added only for the points contained in the new dataframe.

```{r}
# Filter required rows.
midwest_sub <- midwest[midwest$poptotal > 300000, ]
midwest_sub$large_county <- ifelse(midwest_sub$poptotal > 300000, midwest_sub$county, "")

# Base Plot
gg <- ggplot(midwest, aes(x=area, y=poptotal)) + 
  geom_point(aes(col=state, size=popdensity)) + 
  geom_smooth(method="loess", se=F) + xlim(c(0, 0.1)) + ylim(c(0, 500000)) + 
  labs(title="Area Vs Population", y="Population", x="Area", caption="Source: midwest")

# Plot text and label ------------------------------------------------------
gg + geom_text(data=midwest_sub, aes(label=large_county), size=2 ) + labs(subtitle="With ggplot2::geom_text") + theme(legend.position = "None")   # text

gg + geom_label(data=midwest_sub,aes(label=large_county), size=2, alpha=0.25) + labs(subtitle="With ggplot2::geom_label") + theme(legend.position = "None")  # label

# Plot text and label that REPELS eachother (using ggrepel pkg) ------------
library(ggrepel)
gg + geom_text_repel(data=midwest_sub,aes(label=large_county), size=2 ) + labs(subtitle="With ggrepel::geom_text_repel") + theme(legend.position = "None")   # text

gg + geom_label_repel(aes(label=large_county), size=2, data=midwest_sub) + labs(subtitle="With ggrepel::geom_label_repel") + theme(legend.position = "None")   # label


```

## Faceting: Draw multiple plots within one figure

```{r}
data(mpg, package="ggplot2") 

g <- ggplot(mpg, aes(x=displ, y=hwy)) + 
      geom_point() + 
      labs(title="hwy vs displ", caption = "Source: mpg") +
      geom_smooth(method="lm", se=FALSE) + 
      theme_bw()  # apply bw theme
plot(g)
```
#We have a simple chart of highway mileage (hwy) against the engine displacement (displ) for the whole dataset. But what if you want to study how this relationship varies for different classes of vehicles?

## Facet Wrap
#The facet_wrap() is used to break down a large plot into multiple small plots for individual categories. It takes a formula as the main argument. The items to the left of ~ forms the rows while those to the right form the columns.
#By default, all the plots share the same scale in both X and Y axis. You can set them free by setting scales='free' but this way it could be harder to compare between groups.
```{r}

g <- ggplot(mpg, aes(x=displ, y=hwy)) + 
      geom_point() + 
      geom_smooth(method="lm", se=FALSE) + 
      theme_bw()  # apply bw theme

# Facet wrap with common scales
g + facet_wrap( ~ class, nrow=3) + labs(title="hwy vs displ", caption = "Source: mpg", subtitle="Ggplot2 - Faceting - Multiple plots in one figure")  # Shared scales

# Facet wrap with free scales
g + facet_wrap( ~ class, scales = "free") + labs(title="hwy vs displ", caption = "Source: mpg", subtitle="Ggplot2 - Faceting - Multiple plots in one figure with free scales")  # Scales free
```


## Facet Grid
The headings of the middle and bottom rows take up significant space. The facet_grid() would get rid of it and give more area to the charts. The main difference with facet_grid is that it is not possible to choose the number of rows and columns in the grid.

Alright, Let’s create a grid to see how it varies with manufacturer.

```{r}
g <- ggplot(mpg, aes(x=displ, y=hwy)) + 
      geom_point() + 
      labs(title="hwy vs displ", caption = "Source: mpg", subtitle="Ggplot2 - Faceting - Multiple plots in one figure") +
      geom_smooth(method="lm", se=FALSE) + 
      theme_bw()  # apply bw theme

# Add Facet Grid
g1 <- g + facet_grid(manufacturer ~ class)  # manufacturer in rows and class in columns
plot(g1)


#Let’s make one more to vary by cylinder.

g <- ggplot(mpg, aes(x=displ, y=hwy)) + 
      geom_point() + 
      geom_smooth(method="lm", se=FALSE) + 
      labs(title="hwy vs displ", caption = "Source: mpg", subtitle="Ggplot2 - Facet Grid - Multiple plots in one figure") +
      theme_bw()  # apply bw theme

# Add Facet Grid
g2 <- g + facet_grid(cyl ~ class)  # cyl in rows and class in columns.
plot(g2)

##Great!. It is possible to layout both these charts in the sample panel. I prefer the gridExtra() package for this.


library(gridExtra)
gridExtra::grid.arrange(g1, g2, ncol=2)
```
